Models, derivatives, response diversity

### 2024-07-05 models_derivatives_response_diversity.Rmd
rerun with lepomis gibbosus

```{r}
# packages relevant for data analysis
library(tidyverse)
library(here)
library(readr)
library(readxl)
library(gratia)
library(mgcv)
library(forcats)
library(ggpubr)
library(RColorBrewer)
library(broom)
library(grid)
library(DHARMa)
library(mgcViz)
library(gridExtra)
library(grid)

@@

```{r, fig.show='hide'}
# Load total predictions
model_predictions <- readRDS("total_models/df_pred_all.rds")

@@

     saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
  }
}

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i))
      rename(temp = mean_last_7days)

    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
  }
}

@@

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
      fProtocol = factor("VERT"), fLake = factor(random_lake))

@@

    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)
    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))

@@

    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))

@@

# save all the derivatives in one file

df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |> 
  map_dfr(readRDS)

@@

# save total derivatives as RDS (no exclusions)
# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")

@@

# Plot S2: derivatives included in each lake community
overview_derivatives <- all_derivatives |>
  arrange(fLake) |> 
  ggplot(aes(temp, derivative, color = species)) +
  geom_line() +
  facet_wrap(~fLake) +
  theme_bw(base_size = 16) +
  scale_color_viridis(discrete = TRUE, breaks = species_group, option = "G") +
  xlab("Temperature") +
  ylab("Derivatives") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "right",
        legend.title = element_blank()) +
  geom_hline(yintercept = 0, color = "#7A7A7A")

@@

```{r}
# Data of the derivatives before simplifying multiple models
# without changes the 5 species were not run with simplified models. 
all_models_derivatives_no_excl <- readRDS("total_models/df_deriv_all_no_excl.rds")
df_deriv_no_excl <- as_tibble(all_models_derivatives_no_excl)

@@

df_quantiles <- df_deriv_no_excl |> 
  group_by(species, fLake) |> 
  mutate(max_derivative = max(derivative)) |> 
  mutate(mean_derivative = mean(derivative)) |> 
  ungroup()

# calculate quantiles of derivatives
transform(quantile(df_quantiles$derivative,
                   c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))

# add column with the category of quantile each derivative is in
df_quantiles$groups <- cut(df_quantiles$derivative,              
                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
                                1.231212e+07, 1.406101e+07),
                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
                                "95-99%", "99-100%"))

@@

# Plot S3a add lines for the percentiles
plot_percentiles_1 <- plot_percentiles +
  geom_vline(xintercept = 19.9, color = "red") +
  annotate("text", x= 18, y = 0.12, label="90th percentile", angle=90) +
  geom_vline(xintercept = 2.911987e+00, color = "red") +
  annotate("text", x= 1.5, y = 0.12, label="85th percentile", angle=90) +
  theme_classic(base_size = 13) +
  xlab("Mean species derivative per lake") +
  ylab("Percentage") +
 guides(fill = guide_legend(
    title = "Species")) +
  ylim(0, 0.2)

@@

# above 90% = 2.018455e+01
df_quantiles |> 
  filter(derivative > 2.018455e+01) |> 
  distinct(species)

@@

# Breaking mean derivatives in the same percentiles as before
data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
                     breaks = c(-1.579943e+06, -2.342866e+00, -8.822087e-01,
                                -4.543748e-01, 3.454218e-04, 2.485142e-01,
                                5.527792e-01, 2.911987e+00, 2.018455e+01, 1.280898e+02,
                                1.231212e+07, 1.406101e+07),
                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
                                "95-99%", "99-100%"))

@@

# Lepomis is excluded from the analysis
all_deriv <- all_deriv |>
  filter(!species %in% c("Lepomis_gibbosus"))

@@

# Loop to get response diversity measures for each lake 
# loop through list of lakes and calculate response diversity in the lake communities
for (i in lakes_list){

  data <- all_deriv |>
    select(temp, fLake, derivative, species) |> 
    filter(fLake == i)

  df_resp_div <- data |>
    pivot_wider(
      names_from = species,
      values_from = derivative)

@@

  # save response diversity metrics per lake
  saveRDS(df_resp_div, paste0("total_models/lakes_all_models/df_resp_div_", i, ".rds"))

@@

# save the response diversity per lake into one file
resp_div_no_excl <- list.files(path = "total_models/lakes_all_models", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS) |>
  relocate(rdiv, Med, sign, .after = temp)

>>>>

```{r}
# packages relevant for data analysis
library(tidyverse)
library(here)
library(readr)
library(readxl)
library(gratia)
library(mgcv)
library(forcats)
library(ggpubr)
library(RColorBrewer)
library(broom)
library(grid)
library(DHARMa)
library(mgcViz)
library(gridExtra)
library(grid)
library(viridis)

@@

```{r, fig.show='hide'}

# Load total predictions
model_predictions <- readRDS("total_models/df_pred_all.rds")

@@

     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

@@

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

@@

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"), fLake = factor(random_lake))

@@

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

# Model 4: before simplification

species_list <- df_abundance_re |> 
  # exclusion of simplified models
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
                        "Lepomis_gibbosus")) |>
  distinct(Species) |> 
  pull(Species)

    saveRDS(derivatives, paste0("total_models/derivatives_no_simplification/derivatives_", i, "_",  j, ".rds"))

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)
    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

@@

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

@@

# save all the derivatives in one file
# 
df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

@@

# save total derivatives as RDS (no exclusions)
# 
# df_deriv_no_excl <- list.files(path = "total_models/derivatives_no_simplification", pattern = ".rds", full.names = TRUE) |>
#   map_dfr(readRDS)
# 
# saveRDS(df_deriv_no_excl, "total_models/df_deriv_all_no_excl.rds")

@@

# Plot S2: derivatives included in each lake community
overview_derivatives <- all_derivatives |>
  arrange(fLake) |> 
  ggplot(aes(temp, .derivative, color = species)) +
  geom_line() +
  facet_wrap(~fLake) +
  theme_bw(base_size = 16) +
  scale_color_viridis(discrete = TRUE, breaks = species_group, option = "G") +
  xlab("Temperature") +
  ylab("Derivatives") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "right",
        legend.title = element_blank()) +
  geom_hline(yintercept = 0, color = "#7A7A7A")

@@

```{r}
# cyprinus carpio and lepomis should not be simplified

# Data of the derivatives before simplifying multiple models
# without changes the 5 species were not run with simplified models. 
all_models_derivatives_no_excl <- readRDS("total_models/df_deriv_all_no_excl.rds")
df_deriv_no_excl <- as_tibble(all_models_derivatives_no_excl)

@@

df_quantiles <- df_deriv_no_excl |> 
  group_by(species, fLake) |> 
  mutate(max_derivative = max(.derivative)) |> 
  mutate(mean_derivative = mean(.derivative)) |> 
  ungroup()

# calculate quantiles of derivatives
transform(quantile(df_quantiles$.derivative,
                   c(0,0.01,0.05,0.1,0.25,0.5,0.75,0.85,0.9,0.95,0.99,1)))

# add column with the category of quantile each derivative is in
df_quantiles$groups <- cut(df_quantiles$.derivative,              
                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
                                1.303406e+07, 1.406101e+07),
                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
                                "95-99%", "99-100%"))

@@

# Plot S3a add lines for the percentiles
plot_percentiles_1 <- plot_percentiles +
  geom_vline(xintercept = 3.013806e+01, color = "red") +
  annotate("text", x= 26, y = 0.12, label="90th percentile", angle=90) +
  geom_vline(xintercept = 2.018455e+01, color = "red") +
  annotate("text", x= 16, y = 0.12, label="85th percentile", angle=90) +
  theme_classic(base_size = 13) +
  xlab("Mean species derivative per lake") +
  ylab("Percentage") +
 guides(fill = guide_legend(
    title = "Species")) +
  ylim(0, 0.2)

@@

# above 90% = 2.018455e+01
df_quantiles |> 
  filter(.derivative > 2.018455e+01) |> 
  distinct(species)

@@

# Breaking mean derivatives in the same percentiles as before

data_quantiles$percentiles <- cut(data_quantiles$mean_derivative,              
                     breaks = c(-1.579943e+06, -2.706252e+00, -1.269492e+00,
                                -3.348207e-01, 1.844665e-03, 2.262231e-01,
                                9.864616e-01, 2.018455e+01, 3.013806e+01, 1.648215e+06,
                                1.303406e+07, 1.406101e+07),
                     labels = c("0-1%", "1-5%", "5-10%", "10-25%", "25-50%", "50-75%", "75-85%", "85-90%", "90-95%",
                                "95-99%", "99-100%"))

@@

@@

# Loop to get response diversity measures for each lake 
# loop through list of lakes and calculate response diversity in the lake communities

for (i in lakes_list){

  data <- all_deriv |>
    select(.derivative, temp, species, fLake) |> 
    filter(fLake == i)

  df_resp_div <- data |>
    pivot_wider(
      names_from = species,
      values_from = .derivative)

@@

  # save response diversity metrics per lake
  saveRDS(df_resp_div, paste0("total_models/resp_div_lakes/df_resp_div_", i, ".rds"))

@@

# save the response diversity per lake into one file
resp_div_no_excl <- list.files(path = "total_models/resp_div_lakes", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS) |>
  relocate(rdiv, Med, sign, .after = temp)
